Adaptive broadcast tree for multicores

### 2013-04-30 paper.tex
Some more restructuring

@@

\begin{description}
\item[synchrony] Parts of a multicore machine are synchronous. Hence,
  one characteristic of a node is an identifier for the synchrony
  cluster of a node.
\item[$t_{send}$] Time required to send a message, i.e.\ marshalling
  and queuing the message
\item[$t_{recv}$] Time required to receive a message, i.e.\
  dequeuing, unmarshalling and demultiplexing a channel.
\item[failure] Can nodes fail? If so, how? (Byzantine, crash)
\end{description}

@@

% Model given
We assume that the model is given (if it is not, we can do online
measurements and read some specs to find the relevant information). We
just define one ourselves for now or build one manually for a
particular machine. In a real-world scenario, this does not work. The
model needs to be created automatically for a particular machine (due
to the diversity of machines, and also the pace at which hardware changes)

\subsubsection{Example model approximating a 8x4x1 machine}
\label{sec:example_model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Atomic Broadcast}

We build our atomic broadcast based on an overlay network connecting
all nodes of a given multicore machine. First, we discuss what
kind of overlay networks make sense on multicores.

As a first step, we evaluate tree based overlay networks. Other
potentially interesting topologies are rings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tree overlay}

We now look at tree based atomic broadcast implementations.

The first challenge is to find efficient spanning trees for arbitrary
machines. These machines are represented by a model, which we
described earlier.

%-------------------------------------------------
\subsubsection{Divide and conquer}

@@

Another idea is to employ clustering. One example is to base this on
NUMA domains such that every NUMA domain corresponds to one cluster.

%-------------------------------------------------
\subsubsection{Minimum-Spanning Tree}

An important variable in constructing such a tree is the degree of
nodes. If the send cost $t_s$ is smaller than the propagation time
$t_p$,  $d = \lfloor \frac{t_p}{t_s} \rfloor$ messages can be send to
hide propagation of messages. Hence, a tree with degree $d$ should be
ideal to minimize latency, but only for an acknowledged protocol.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rings}

\stefan{Just a random collection of ideas}

%-------------------------------------------------
\subsubsection{Load balancing in a tree}

We could build several rings, that do not always include all
coordinators and work with read- and write-sets. We can select these
rings such that they nicely integrate with the physical topology. They
can also have different sizes (e.g.\ a smaller one for read operations
and larger ones for writes).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Example tree}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Broadcast protocol}

% Our implementation
Assuming the tree, we now build an atomic broadcast for multicore
machines. Our implementation is based on the Barrelfish UMP
interconnect driver. It provides reliable channels with in-order
delivery. Communication starts at the root node, which acts as a
sequentializer. See~\ref{algo:ab} for details.

@@

\end{algorithm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barrelfish implementation}

When implementing the protocol in Barrelfish, we found several
practical problems, which we will discuss briefly in this section.

We kick-start our protocol connecting every process with all other
processes to get a fully-meshed network of channels.

We use a round-based algorithm to open channels. Then, every node
knows exactly the source of an incoming connection. We formalize this
algorithm in Algorithm~\ref{algo:ab_bind}.

This is required for Barrelfish UMP communication channels since on
bind, no source identifier is send along\footnote{Check if this is
  actually true, and even if it is true, if is a Barrelfish problem,
  and not a general one}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mechanisms}

\stefan{This section feels out of place. Rewrite it and put it
  somewhere else}

\subsection{Aggregation for convergecast}

Convergecast is a way of collecting information from nodes (i.e.\ it
is kind of the reverse of a broadcast). On the return path,
information can be aggregated. One example is batching, where several
pieces of information are grouped in the same message to reduce packet
processing overhead. Another example is aggregation, i.e.\
pre-processing of data. One example is pre-calculating the average
value of child messages if the sink node is not interested in
individual values.

Similarly to what has been done in wireless sensor networks (where it
also matters to reduce the number of messages, but for other reasons:
power consumption), we can do aggregation in nodes. In difference to
traditional distributed systems, this works, because it is easy to
deploy custom software on every node in the network. Classical
distributed systems do not typically allow this. Furthermore, reducing
the number of messages at the price of higher complexity does not make
sense in classical systems.

Examples for aggregation: number of nodes agreeing to something, find
capabilities (concatenate core ids).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example}

\paragraph{Observations} Reducing the number of children also helps in
select. It reduces the number of ``channels'' to poll, and therefore
the latency of detecting messages. Instead of

@@

This section shows results for a simulation based on the model
introduced in Section~\ref{sec:example_model}

The time given is a fictional time unit at this point.

\begin{table}[htb]
  \centering
  \begin{tabular}{llll}
    \toprule
    overlay & scheduling & time & factor \\
    \midrule
    MST & naive         & 241 & 1.00 \\
    MST & longest first & 151 & 0.63 \\
    \bottomrule
  \end{tabular}
  \caption{Simulation results for simplified machine model}
  \label{tab:sim_results}
\end{table}

%-------------------------------------------------
\subsection{Real hardware}

@@

\paragraph{Minimal cost of sending a tree-broadcast} %
Plot~\ref{pgfplot:201303141819} shows the cost of flooding a sub-tree
(with the 50\% worst measurements dropped). The group communication is
based on a binary tree. Core 0 is the root, cores 1 and 2 are its
children etc.

The average cost is really high (probably a scheduling issue). But the
minimal numbers show what is possible. The numbers achieved are easily
explainable. The cost is increasing logarithmic with the number of
nodes reached by the broadcast (as expected). Every level in the tree
adds an additional 3500 cycles to the tree. Node 0 takes significantly
longer. I don't know yet why that is.

@@

\input{gruyere_bc_seq}

\begin{itemize}
\item Move server to a core != 0
\item Use low-level UMP stuff
{
\renewcommand{\labelitemi}{\checkmark}
\item disable yielding
}
\item run several request in parallel, give them IDs, can use in the
  measurement struct. Good for everything having transactions (DB,
  transactional memory, consistency in replication systems)
\end{itemize}

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\end{itemize}

@@

Octopus is a coordination service implementation for the Barrelfish
Multikernel operating system. It is, however, centralized. The idea is
to make Octopus a decentralized application.

@@

The question is whether a key-value store as a backend for
synchronization primitives even makes sense. As we want to show
efficiency, we probably rather want to implement new synchronization
algorithms from scratch.

@@

@@

\begin{description}
\item[Updates on global state] Slaves direct their updates to the
  local coordinator. Coordinators send the update request $r$ along
  the ring and wait for them to be send back. While passing on
  requests, coordinators can set a flag to indicate conflicts. A
  conflict is another colliding request $r_o$ that is already in
  flight. I think we will have to send activity vectors along with
  this information! Otherwise, the coordinators requesting $r$ and
  $r_o$ will both prevent the other thing from happening and none of
  them will be applied! Maybe we need a two-phase commit.
\item[Barriers] The coordinator knows how many of its slaves need to
  enter the barrier. It can then just aggregate local barrier\_enters
  calls and only communicate this information to the others once all
  children are ready. 
\end{description}

>>>>

\newcommand{\etal}{{\it et al.}\xspace}

@@

\begin{description}
\item[synchrony] Parts of a multicore machine are synchronous. Hence,
  one characteristic of a node is an identifier for the synchrony
  cluster of a node.
\item[send time $t_{send}$] Time required to send a message, i.e.\ marshalling
  and queuing the message
\item[receive $t_{recv}$] Time required to receive a message, i.e.\
  dequeuing, unmarshalling and demultiplexing a channel.
\item[failure] Can nodes fail? If so, how? (Byzantine, crash)
\end{description}

@@

% Model given
We assume that the model is given. We just define one ourselves for
now or build one manually for a particular machine. In a real-world
scenario, this does not work. The model needs to be created
automatically for a particular machine (due to the diversity of
machines, and also the pace at which hardware changes)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Atomic Broadcast for Multicores}

For a specific model, we now need to find ``the right'' implementation
of algorithms to program it. We build our atomic broadcast based on a
subset of links offered by the model such that all nodes of a given
multicore machine can be reached. Atomic broadcast guarantees that 1)
if one node receives a message, all other nodes receive it as well,
and 2) that all nodes see the message in the same order.

First, we discuss what kind of topologies we use to communicate
between nodes. We evaluate tree based implementations. Other
potentially interesting topologies are rings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tree Topology}

We now look at tree-based atomic broadcast implementations. The first
challenge is to find efficient spanning trees for arbitrary
machines. We now discuss different strategies to select trees based on
the model representing the machine.

%-------------------------------------------------
\subsubsection{Minimum-Spanning Tree}
\label{sec:mst_tree}

An important variable in constructing such a tree is the degree of
nodes. If the send cost $t_s$ is smaller than the propagation time
$t_p$, $d = \lfloor \frac{t_p}{t_s} \rfloor$ messages can be send to
hide propagation of messages. Hence, a tree with degree $d$ should be
ideal to minimize latency. For multicore interconnect networks,
however, we expect $d<1$. Thus, we do not look into this in detail
right now.

Reducing the degree of nodes also helps for the select operation. It
reduces the number of memory locations to poll, and therefore the
latency of detecting messages.

We discussed earlier in Section~\ref{sec:model_hierarchy} that
multicore machines are hierarchical.%

@@

Another idea is to employ clustering on NUMA domains such that every
NUMA domain corresponds to one cluster.

%-------------------------------------------------
\subsubsection{Example tree}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation}

% Our implementation
Assuming the tree, we now implement an atomic broadcast for multicore
machines. Our implementation is based on the Barrelfish UMP
interconnect driver. It provides reliable channels with in-order
delivery. Flow control is build in for reliability. Communication
starts at the root node, which acts as a
sequentializer. See~\ref{algo:ab} for details.

@@

\end{algorithm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Issues with the Barrelfish implementation}

When implementing the protocol in Barrelfish, we found several
practical problems, which we will discuss briefly in this section.
We kick-start our protocol connecting every process with all other
processes to get a fully-meshed network of channels.
We use a round-based algorithm to open channels. Then, every node
knows exactly the source of an incoming connection. We formalize this
algorithm in Algorithm~\ref{algo:ab_bind}.
This is required for Barrelfish UMP communication channels since on
bind, no source identifier is send along\footnote{Check if this is
  actually true, and even if it is true, if is a Barrelfish problem,
  and not a general one}.

@@

@@

This section shows results for a simulation based on the model
introduced in Section~\ref{sec:mst_tree}. The time given is a
fictional time unit at this point.

\begin{table}[htb]
  \centering
  \begin{tabular}{llll}
    \toprule
    topology & scheduling & time & factor \\
    \midrule
    MST & naive         & 241 & 1.00 \\
    MST & longest first & 151 & 0.63 \\
    \bottomrule
  \end{tabular}
  \caption{Simulation results for simplified machine model}
  \label{tab:sim_results}
\end{table}

%-------------------------------------------------
\subsubsection{Real hardware}

@@

Plot~\ref{pgfplot:201303141819} shows the cost of flooding a sub-tree
(with the 50\% worst measurements dropped). The group communication is
based on a binary tree. Core 0 is the root, cores 1 and 2 are its
children etc.

The average cost is really high (probably a scheduling issue). But the
minimal numbers show what is possible. The numbers achieved are easily
explainable. The cost is increasing logarithmic with the number of
nodes reached by the broadcast (as expected). Every level in the tree
adds an additional 2500 to 3000 cycles to the tree. Node 0 takes
significantly longer. I don't know yet why that is.

@@

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\item Try other topologies (Frans Kaashoek), e.g.\ rings
\item \emph{Load balancing in a tree} We could build several rings,
  that do not always include all coordinators and work with read- and
  write-sets. We can select these rings such that they nicely
  integrate with the physical topology. They can also have different
  sizes (e.g.\ a smaller one for read operations and larger ones for
  writes).
\item \emph{Aggregation for convergecast} %
  Convergecast is a way of collecting information from nodes (i.e.\ it
  is kind of the reverse of a broadcast). On the return path,
  information can be aggregated. One example is batching, where several
  pieces of information are grouped in the same message to reduce packet
  processing overhead. Another example is aggregation, i.e.\
  pre-processing of data. One example is pre-calculating the average
  value of child messages if the sink node is not interested in
  individual values. %
  Similarly to what has been done in wireless sensor networks (where it
  also matters to reduce the number of messages, but for other reasons:
  power consumption), we can do aggregation in nodes. In difference to
  traditional distributed systems, this works, because it is easy to
  deploy custom software on every node in the network. Classical
  distributed systems do not typically allow this. Furthermore, reducing
  the number of messages at the price of higher complexity does not make
  sense in classical systems. %
  Examples for aggregation: number of nodes agreeing to something, find
  capabilities (concatenate core ids). 
\end{itemize}

@@

Octopus is a coordination service implementation for the Barrelfish
Multikernel operating system based on a key-value store
data-backend. It is, however, centralized. The idea is to make Octopus
a decentralized application.

@@

@@

Slaves direct their updates to the local coordinator. Coordinators
send the update request $r$ along the ring and wait for them to be
send back. While passing on requests, coordinators can set a flag to
indicate conflicts. A conflict is another colliding request $r_o$ that
is already in flight. I think we will have to send activity vectors
along with this information! Otherwise, the coordinators requesting
$r$ and $r_o$ will both prevent the other thing from happening and
none of them will be applied! Maybe we need a two-phase commit.

@@

%-------------------------------------------------
\subsubsection{Synchronization}

It is doubtful that a key-value store as a backend for synchronization
primitives makes sense. As we want to show efficiency, we probably
rather want to implement new synchronization algorithms from scratch.

The coordinator knows how many of its slaves need to enter the
barrier. It can then just aggregate local barrier\_enters calls and
only communicate this information to the others once all children are
ready. Local communication can be shared-memory based, while global
communication should probably rather be message-passing based.

We will have to compare the performance of such an implementation with
a purely shared-memory based implementation like pthreads and
something entirely message-passing based like MPI versions.
